Ecosystem carbon balance in the Hawaiian Islands under different scenarios of future climate and land use change

### 2021-03-23 Manuscript/HI_Model_ms.Rmd
updated manuscript files

```{r setup, include=FALSE}
options(tinytex.verbose = TRUE)
knitr::opts_chunk$set(echo = FALSE)
```
\centering
Paul C. Selmants^1,6^, Benjamin M. Sleeter^2^,  Jinxun Liu^1^, Tamara S. Wilson^1^,

Clay Trauernicht^3^, Abby G. Frazier^4^, Gregory P. Asner^5^

@@

**Affiliations:**  
^1^U.S. Geological Survey, Moffett Field, CA, USA     
^2^U.S. Geological Survey, Seattle, WA, USA    
^3^University of Hawaiʻi at Mānoa, Honolulu, HI, USA  
^4^The East-West Center, Honolulu, HI, USA    
^5^Arizona State University, Tempe, AZ, USA    
^6^Author to whom correspondence should be addressed

@@

The State of Hawaiʻi exemplifies the challenges associated with projecting the interactive effects of future climate and land use change on ecosystem carbon balance at a regional scale. Hawaiʻi was the first U.S. state to enact legislation committing to full carbon neutrality, requiring the state to account for and offset all of its greenhouse gas emissions by 2045 (State of Hawaiʻi Acts 15 and 16). This legislation emphasizes the mitigation potential of natural ecosystems as a key component to emissions reduction, necessitating baseline estimates and future projections of land carbon sink strength. However, Hawaiʻi's challenging terrain complicates these assessment efforts. The main Hawaiian Islands are a complex mosaic of natural and human-dominated landscapes overlain by steep climate gradients across relatively short distances (supplmental figure 3), with mean annual temperature ranging from ~4-24° C [@giambelluca_evapotranspiration_2014] and mean annual rainfall ranging from ~180-9500 mm [@giambelluca_online_2013]. Temperatures have risen rapidly in the Hawaiian Islands since the mid 1970s [@giambelluca_secular_2008] and a long-term drying trend has persisted since the early 1920s [@frazier_spatial_2017], resulting in reduced forest biomass and productivity [@barbosa_effects_2017]. These same drying and warming trends have increased the frequency and intensity of wildland fire [@trauernicht_contemporary_2015; @trauernicht_vegetationrainfall_2019] with predictable negative effects on ecosystem carbon balance [@selmants_baseline_2017]. Ecosystem carbon stocks across the main Hawaiian Islands have also been strongly influenced by the legacy of past land use change [@osher_effect_2003; @asner_high-resolution_2011]. Thousands of hectares of land were deforested beginning in the late 19th century to clear land for sugar plantations and cattle pasture [@cuddihy_alteration_1990]. Since the mid-20th century, much of this agricultural land has been steadily converted to urban areas, commercial forestry plantations, or simply abandoned and colonized by non-native grass species [@suryanata_diversified_2009; @perroy_evolving_2016]. Although these past trends surely inform the future impact of climate and land use change on ecosystem carbon balance, high spatial and temporal heterogeneity complicates realistic projection efforts. To date only one study has attempted to integrate land use, climate, and natural disturbances into future projections of Hawaiian ecosystem carbon balance, with projections limited to the mid-21st century under a single land use change scenario and moderate radiative forcing [SRES A1B, equivalent to RCP 6; @selmants_baseline_2017].

@@

 Initial carbon stocks and baseline carbon flows were estimated based on the moisture zone (supplmental figure 1), state type, and age of each simulation cell using a lookup table derived from the Integrated Biosphere Simulator [IBIS; @foley_integrated_1996; @liu_critical_2020], a process-based dynamic global vegetation model. We initiated IBIS with minimal vegetation and simulated forward for 110 years using 30-year climate normals for the Hawaiian Islands [@giambelluca_online_2013; @giambelluca_evapotranspiration_2014]. We calibrated IBIS carbon stocks with statewide gridded datasets of soil organic carbon [@ssurgo_2016] and forest aboveground live biomass [@asner_rapid_2016]. We also calibrated gross photosynthesis in IBIS using a Hawaiʻi-specific gridded dataset derived from MODIS satellite imagery [@kimball_evaluating_2017].

 Net primary production for each simulation cell was calculated as the mean IBIS-derived value for each combination of moisture zone and state type adjusted with a spatially explicit stationary growth multiplier to reflect local variation driven by microclimate [@sleeter_effects_2019]. We calculated this spatial growth multiplier as the NPP anomaly for each simulation cell relative to mean NPP values for each combination of moisture zone and state type based on empirical relationships between total annual NPP and mean annual rainfall or temperature [@schuur_productivity_2003; @del_grosso_global_2008] using Hawaiʻi-specific climate data [@giambelluca_online_2013; @giambelluca_evapotranspiration_2014]. Soil carbon flux to the atmosphere (R~h~) and aquatic soil carbon losses (leaching and overland flow) were estimated as the ratio of the IBIS-derived flux for each combination of moisture zone and state type to the microclimate-adjusted NPP value for each simulation cell. All other carbon flow rates were estimated as the ratio of the mean IBIS-derived flux for each combination of moisture zone and state type to the size of the originating carbon stock at each age [@sleeter_effects_2018; @daniel_integrating_2018]. Climate change impacts on carbon flows were represented by temporal growth and decay multipliers applied to each simulation cell based on statistically downscaled CMIP5 climate projections for the Hawaiian Islands under each of two radiative forcing scenarios [RCP 4.5 and RCP 8.5; @timm_statistical_2015; @timm_future_2017]. The impact of future changes in rainfall and temperature on NPP were represented by annual growth multipliers calculated using empirical NPP models [@schuur_productivity_2003; @del_grosso_global_2008] and climate model projections of temperature and rainfall for each radiative forcing scenario. The effect of future warming on turnover rates of dead organic matter were represented by temporal decay multipliers calculated using Q10 functions and climate model temperature projections for each radiative forcing scenario. We applied a Q10 of 2.0 for wood and soil organic matter decay flows [@kurz_cbm-cfs3:_2009; @sleeter_effects_2019] and a Q10 of 2.17 for litter decay flows [@bothwell_leaf_2014]. Transition-triggered carbon flows resulting from disturbances associated with land use change, timber harvesting, and wildfire were based on values from @don_impact_2011, @selmants_baseline_2017, and @daniel_integrating_2018.

@@

```{r fig.align = 'center', out.width = "99%", fig.cap = "Projected changes in total ecosystem carbon storage (a) and individual carbon stocks (b) for the seven main Hawaiian Islands. Solid lines indicate the mean of 30 Monte Carlo realizations for each scenario, with shaded areas indicating the minimum and maximum range of Monte Carlo realizations."}
knitr::include_graphics("./fig_images/fig2_Cstocks.png")
```
Net primary production (NPP) for the seven main Hawaiian Islands declined across all four scenarios, driven primarily by climate change and to a lesser extent by land use change (Fig. 3). The combination of high radiative forcing (RCP 8.5) and high land use change led to the steepest decline in NPP over time, driven by intense long-term drying on the leeward sides of islands under RCP 8.5 (supplemental figure 4) and sustained losses of forest and shrubland land area in the high land use scenario (supplmental figure 5). In contrast, climate change led to increased heterotrophic respiration (R~h~) over time, such that more intense warming under RCP 8.5 (supplmental figure 4) resulted in R~h~ being ~ 3% higher by 2100 than under RCP 4.5 (figure 3). Heterotrophic respiration declined substantially over time in the high land use scenario (figure 3) because of long-term reductions in forest and shrubland land area (supplemental figure 5), similar to trends in NPP. Transition-triggered carbon fluxes to the atmosphere from land use, land use change, and wildfire were largely independent of changes in climate, stabilizing by mid-century at an average of ~0.4 Tg y^-1^ in the high land use scenario and ~0.2 Tg y^-1^ in the low land use scenario (figure 3). Uncertainty around transition-triggered carbon fluxes were higher in the high land use scenario, driven primarily by greater variability in wildland fire probabilities.

@@

This study was funded by the U.S. Geological Survey Biological Carbon Sequestration Program. Thanks to Leonardo Frid and Colin Daniel of ApexRMS for assistance with SyncroSim software, and to Nicholas Koch of Forest Solutions, Inc. for information on _Eucalyptus_ spp. harvesting in Hawaiʻi. Thanks also to Christian Giardina and Zhiliang Zhu for providing the initial impetus for this research. Any use of trade, firm, or product names is for descriptive purposes only and does not imply endorsement by the U.S. Government.

>>>>

```{r setup, include=FALSE}
options(tinytex.verbose = TRUE)
knitr::opts_chunk$set(echo = FALSE)
```
\centering
Paul C. Selmants^1,7^, Benjamin M. Sleeter^2^,  Jinxun Liu^1^, Tamara S. Wilson^1^,

Clay Trauernicht^3^, Abby G. Frazier^4,5^, Gregory P. Asner^6^

@@

**Affiliations:**  
^1^U.S. Geological Survey, Moffett Field, CA, USA     
^2^U.S. Geological Survey, Seattle, WA, USA    
^3^University of Hawaiʻi at Mānoa, Honolulu, HI, USA  
^4^East-West Center, Honolulu, HI, USA 
^5^Clark University, Worcester, MA, USA   
^6^Arizona State University, Tempe, AZ, USA    
^7^Author to whom correspondence should be addressed

@@

The State of Hawaiʻi exemplifies the challenges associated with projecting the interactive effects of future climate and land use change on ecosystem carbon balance at a regional scale. Hawaiʻi was the first U.S. state to enact legislation committing to full carbon neutrality, requiring the state to account for and offset all of its greenhouse gas emissions by 2045 (State of Hawaiʻi Acts 15 and 16). This legislation emphasizes the mitigation potential of natural ecosystems as a key component to emissions reduction, necessitating baseline estimates and future projections of land carbon sink strength. However, Hawaiʻi's challenging terrain complicates these assessment efforts. The main Hawaiian Islands are a complex mosaic of natural and human-dominated landscapes overlain by steep climate gradients across relatively short distances, with mean annual temperature ranging from ~4-24° C [@giambelluca_evapotranspiration_2014] and mean annual rainfall ranging from ~200-10,200 mm [@giambelluca_online_2013]. Temperatures have risen rapidly in the Hawaiian Islands since the mid 1970s [@giambelluca_secular_2008; @mckenzie_temperature_2019] and a long-term drying trend has persisted since the early 1920s [@frazier_spatial_2017], resulting in reduced forest biomass and productivity [@barbosa_effects_2017]. These same drying and warming trends have increased the frequency and intensity of wildland fire [@trauernicht_contemporary_2015; @trauernicht_vegetationrainfall_2019] with predictable negative effects on ecosystem carbon balance [@selmants_baseline_2017]. Ecosystem carbon stocks across the main Hawaiian Islands have also been strongly influenced by the legacy of past land use change [@osher_effect_2003; @asner_high-resolution_2011]. Thousands of hectares were deforested beginning in the late 19th century to clear land for sugar plantations and cattle pasture [@cuddihy_alteration_1990]. Since the mid-20th century, much of this agricultural land has been steadily converted to urban areas, commercial forestry plantations, or simply abandoned and colonized by non-native grass species [@suryanata_diversified_2009; @perroy_evolving_2016]. Although these past trends surely inform the future impact of climate and land use change on ecosystem carbon balance, high spatial and temporal heterogeneity complicates realistic projection efforts. To date only one study has attempted to integrate land use, climate, and natural disturbances into future projections of Hawaiian ecosystem carbon balance, with projections limited to the mid-21st century under a single land use change scenario and moderate radiative forcing [SRES A1B, equivalent to RCP 6; @selmants_baseline_2017].

@@

 Initial carbon stocks and baseline carbon flows were estimated based on the moisture zone (supplemental figure 1), state type, and age of each simulation cell using a lookup table derived from the Integrated Biosphere Simulator [IBIS; @foley_integrated_1996; @liu_critical_2020], a process-based dynamic global vegetation model. We initiated IBIS with minimal vegetation and simulated forward for 110 years using 30-year climate normals for the Hawaiian Islands [@giambelluca_online_2013; @giambelluca_evapotranspiration_2014]. We calibrated IBIS carbon stocks with statewide gridded datasets of soil organic carbon [@ssurgo_2016] and forest aboveground live biomass [@asner_rapid_2016]. We also calibrated gross photosynthesis in IBIS using a Hawaiʻi-specific gridded dataset derived from MODIS satellite imagery [@kimball_evaluating_2017].

 Net primary production for each simulation cell was calculated as the mean IBIS-derived value for each combination of moisture zone and state type adjusted with a spatially explicit stationary growth multiplier to reflect local variation driven by microclimate [@sleeter_effects_2019]. We calculated this spatial growth multiplier as the NPP anomaly for each simulation cell relative to mean NPP values for each combination of moisture zone and state type based on empirical relationships between total annual NPP and mean annual rainfall or temperature [@schuur_productivity_2003; @del_grosso_global_2008] using Hawaiʻi-specific climate data [@giambelluca_online_2013; @giambelluca_evapotranspiration_2014]. Soil carbon flux to the atmosphere (R~h~) and aquatic soil carbon losses (leaching and overland flow) were estimated as the ratio of the IBIS-derived flux for each combination of moisture zone and state type to the microclimate-adjusted NPP value for each simulation cell. All other carbon flow rates were estimated as the ratio of the mean IBIS-derived flux for each combination of moisture zone and state type to the size of the originating carbon stock at each age [@sleeter_effects_2018; @daniel_integrating_2018]. Climate change impacts on carbon flows were represented by temporal growth and decay multipliers applied to each simulation cell based on mid-century (2049-2069) and end-of-century (2070-2099) statistically downscaled CMIP5 temperature and rainfall projections for the Hawaiian Islands under each of two radiative forcing scenarios [RCP 4.5 and RCP 8.5; @timm_statistical_2015; @timm_future_2017]. The impact of future changes in rainfall and temperature on NPP were represented by annual growth multipliers calculated using empirical NPP models [@schuur_productivity_2003; @del_grosso_global_2008] and climate model projections of temperature and rainfall for each radiative forcing scenario. The effect of future warming on turnover rates of dead organic matter were represented by temporal decay multipliers calculated using Q10 functions and climate model temperature projections for each radiative forcing scenario. We applied a Q10 of 2.0 for wood and soil organic matter decay flows [@kurz_cbm-cfs3:_2009; @sleeter_effects_2019] and a Q10 of 2.17 for litter decay flows [@bothwell_leaf_2014]. Transition-triggered carbon flows resulting from disturbances associated with land use change, timber harvesting, and wildfire were based on values from @don_impact_2011, @selmants_baseline_2017, and @daniel_integrating_2018.

@@

```{r fig.align = 'center', out.width = "99%", fig.cap = "Projected changes in total ecosystem carbon storage (a) and individual carbon stocks (b) for the seven main Hawaiian Islands. Solid lines indicate the mean of 30 Monte Carlo realizations for each scenario, with shaded areas indicating the minimum and maximum range of Monte Carlo realizations."}
knitr::include_graphics("./fig_images/fig2_Cstocks.png")
```
Net primary production (NPP) for the seven main Hawaiian Islands declined across all four scenarios, driven primarily by climate change and to a lesser extent by land use change (Fig. 3). The combination of high radiative forcing (RCP 8.5) and high land use change led to the steepest decline in NPP over time, driven by intense long-term drying on the leeward sides of islands under RCP 8.5 (supplemental figure 4) and sustained losses of forest and shrubland land area in the high land use scenario (supplemental figure 5). In contrast, climate change led to increased heterotrophic respiration (R~h~) over time, such that more intense warming under RCP 8.5 (supplemental figure 4) resulted in R~h~ being ~ 3% higher by 2100 than under RCP 4.5 (figure 3). Heterotrophic respiration declined substantially over time in the high land use scenario (figure 3) because of long-term reductions in forest and shrubland land area (supplemental figure 5), similar to trends in NPP. Transition-triggered carbon fluxes to the atmosphere from land use, land use change, and wildfire were largely independent of changes in climate, stabilizing by mid-century at an average of ~0.4 Tg y^-1^ in the high land use scenario and ~0.2 Tg y^-1^ in the low land use scenario (figure 3). Uncertainty around transition-triggered carbon fluxes were higher in the high land use scenario, driven primarily by greater variability in wildland fire probabilities.

@@

This study was funded by the U.S. Geological Survey Biological Carbon Sequestration Program and the Pacific Islands Climate Adaptation Science Center. Thanks to Leonardo Frid and Colin Daniel of ApexRMS for assistance with SyncroSim software, and to Nicholas Koch of Forest Solutions, Inc. for information on _Eucalyptus_ spp. harvesting in Hawaiʻi. Thanks also to Christian Giardina and Zhiliang Zhu for providing the initial impetus for this research. Any use of trade, firm, or product names is for descriptive purposes only and does not imply endorsement by the U.S. Government.
